### 2018-05-18 paper/intro.tex
Do some copyediting and typesetting

\section{Introduction}

In the two decades since they were introduced by \citet{wadler89}, \emph{type classes} have proven a versatile apparatus for structuring programs. Consider a class intended to characterize types $\tVar$ for which a monoid can be defined:
\begin{flalign*}
  & \eClass{\tMonoid{\tVar}} & \\
  & \quad \tAnno{\eIdentity}{\tVar} & \\
  & \quad \tAnno{\eCombine}{\tArrow{\tVar}{\tArrow{\tVar}{\tVar}}} &
\end{flalign*}
Programmers can use this class to define polymorphic functions over all monoids:
\begin{flalign*}
  & \tAnno{\eSquare}{\tMonoid{\tVar} \Rightarrow \tArrow{\tVar}{\tVar}} & \\
  & \eApp{\eSquare}{\eX} = \eApp{\eApp{\eCombine}{\eX}}{\eX} &
\end{flalign*}
Finally, an \emph{instance} of the class can be provided for a specific type:
\begin{flalign*}
  & \eInstance{\tMonoid{\integer}} & \\
  & \quad \eIdentity = \numberName{1} & \\
  & \quad \eCombine = \eMultiply &
\end{flalign*}
where $\tAnno{\numberName{1}}{\integer}$ and $\tAnno{\eMultiply}{\tArrow{\integer}{\tArrow{\integer}{\integer}}}$. The instance is a witness to justify the well-definedness of expressions like $\eApp{\eSquare}{\numberName{3}}$.

More recently, algebraic effects \citep{plotkin03} and handlers \citep{plotkin09} have become an attractive alternative to monads for modeling computational effects due to their compositionality.

### 2018-05-18 scripts/lint-tex.rb

@@ -71,6 +71,7 @@ PRELUDE_ARITIES = {
   'end' => [1],
   'endAcm' => [0],
   'endNonAcm' => [0],
+  'enspace' => [0],
   'fbox' => [1],
   'fboxsep' => [0],
   'fi' => [0],

### 2018-05-19 implementation/src/Evaluation.hs
Add kinds

@@ -15,11 +15,11 @@ eval (FEApp e1 e2) = do
   case e3 of
     FEAbs x _ e5 -> eval $ subst x e4 e5
     _ -> throwError $ "Cannot apply " ++ show e3 ++ " to " ++ show e4
-eval (FETAbs a e) = return $ FETAbs a e
+eval (FETAbs a k e) = return $ FETAbs a k e
 eval (FETApp e1 t) = do
   e2 <- eval e1
   case e2 of
-    FETAbs a e3 -> eval $ subst a t e3
+    FETAbs a _ e3 -> eval $ subst a t e3
     _ -> throwError $ "Cannot apply " ++ show e2 ++ " to " ++ show t
 eval FETrue = return FETrue
 eval FEFalse = return FEFalse

### 2018-05-19 implementation/src/Syntax.hs

@@ -8,9 +8,12 @@ module Syntax
   , EVarName(..)
   , FTerm(..)
   , FreeEVars
+  , FreeKVars
   , FreeTCons
   , FreeTVars
   , ITerm(..)
+  , KVarName(..)
+  , Kind(..)
   , Subst
   , TConName(..)
   , TVarName(..)
@@ -21,6 +24,7 @@ module Syntax
   , boolType
   , collectBinders
   , freeEVars
+  , freeKVars
   , freeTCons
   , freeTVars
   , intName
@@ -62,6 +66,15 @@ instance Show TConName where
   show (UserTConName s) = s
   show (AutoTConName i) = "%" ++ show i
 
+data KVarName
+  = UserKVarName String -- Can be referred to in source programs
+  | AutoKVarName Integer -- Cannot be referred to in source programs
+  deriving (Eq, Ord)
+
+instance Show KVarName where
+  show (UserKVarName s) = s
+  show (AutoKVarName i) = "@" ++ show i
+
 data ITerm
   = IEVar EVarName
   | IEAbs EVarName
@@ -100,6 +113,7 @@ data FTerm
   | FEApp FTerm
           FTerm
   | FETAbs TVarName
+           Kind
            FTerm
   | FETApp FTerm
            Type
@@ -126,7 +140,12 @@ data Type
   | TCon TConName
          [Type]
   | TForAll TVarName
+            Kind
             Type
+
+data Kind
+  = KVar KVarName
+  | KType
 
 -- Built-in type constructors
 boolName :: TConName
@@ -166,6 +185,9 @@ class FreeTVars a where
 class FreeTCons a where
   freeTCons :: a -> [TConName]
 
+class FreeKVars a where
+  freeKVars :: a -> [KVarName]
+
 instance FreeEVars ITerm where
   freeEVars (IEVar x) = [x]
   freeEVars (IEAbs x _ e) = filter (/= x) (freeEVars e)
@@ -189,6 +211,9 @@ instance FreeEVars ITerm where
 -- variables of an ITerm come from type annotations, but free type variables in
 -- annotations are interpreted as implicitly existentially bound (i.e., they
 -- aren't really free).
+--
[…]

### 2018-05-19 implementation/src/Inference.hs
Replace missing type annotations with existential variables

@@ -226,10 +226,7 @@ infer (IEVar x) = do
     Just t -> return (FEVar x, t, emptySubst)
     Nothing -> throwError $ "Undefined variable: " ++ show x
 infer (IEAbs x t1 e1) = do
-  t2 <-
-    case t1 of
-      Just t2 -> sanitizeAnnotation t2
-      Nothing -> TVar <$> freshTVar
+  t2 <- sanitizeAnnotation t1
   (e2, t3, theta) <-
     withUserEVar x t2 $ do
       (e2, t3, theta) <- infer e1

### 2018-05-20 paper/intro.tex
Use the LaTeX (rather than the Tex) syntax for inline math

@@
 In the two decades since they were introduced by \citet{wadler89}, \emph{type classes} have proven a versatile apparatus for structuring programs.
-Consider a class intended to characterize types $\tVar$ for which a monoid can be defined:
+Consider a class intended to characterize types \(\tVar\) for which a monoid can be defined:
 \begin{flalign*}
@@
 \end{flalign*}
-where $\tAnno{\numberName{1}}{\integer}$ and $\tAnno{\eMultiply}{\tArrow{\integer}{\tArrow{\integer}{\integer}}}$.
-The instance is a witness to justify the well-definedness of expressions like $\eApp{\eSquare}{\numberName{3}}$.
+where \(\tAnno{\numberName{1}}{\integer}\) and \(\tAnno{\eMultiply}{\tArrow{\integer}{\tArrow{\integer}{\integer}}}\).
+The instance is a witness to justify the well-definedness of expressions like \(\eApp{\eSquare}{\numberName{3}}\).
 

### 2018-05-23 implementation/src/Inference.hs
Flip the order of state and exception in the TypeCheck monad transformer stack

@@ -12,7 +12,8 @@ module Inference
 
 import Control.Monad (foldM, when)
 import Control.Monad.Except (ExceptT, runExceptT, throwError)
-import Control.Monad.State (State, evalState, get, put)
+import Control.Monad.State (StateT, evalStateT, get, put)
+import Data.Functor.Identity (Identity, runIdentity)
 import Data.List (nub)
 import Data.Map (Map)
 import qualified Data.Map as Map
@@ -55,10 +56,7 @@ import Syntax
 -- 1. The ability to read and update the context (also via State)
 -- 2. The ability to throw errors (via ExceptT)
 type TypeCheck
-   = ExceptT String (State ( Integer
-                           , Map EVarName Type
-                           , Map TVarName Kind
-                           , Map TConName Kind))
+   = StateT (Integer, Map EVarName Type, Map TVarName Kind, Map TConName Kind) (ExceptT String Identity)
 
 -- Add a term variable to the context.
 withUserEVar :: EVarName -> Type -> TypeCheck a -> TypeCheck a
@@ -506,8 +504,10 @@ simplify (FEConcat e1 e2) = FEConcat (simplify e1) (simplify e2)
 typeCheck :: ITerm -> Either String (FTerm, Type)
 typeCheck e1 =
   let result =
-        evalState
-          (runExceptT (infer e1))
+        runIdentity $
+        runExceptT $
+        evalStateT
+          (infer e1)
           ( 0
           , Map.empty
           , Map.empty

### 2018-05-24 implementation/src/Substitution.hs
Add a substitution accumulator in the context

@@ -85,7 +85,7 @@ instance SingletonSubst KVarName Kind where
     idempotencyCheck $ Substitution Map.empty (Map.singleton b k)
 
 -- Compose two substitutions. The substitutions are in diagrammatic order, that
--- is, theta2 comes after theta1.
+-- is, the second substitution comes after the first.
 composeSubst :: Substitution -> Substitution -> Substitution
 composeSubst (Substitution m1 n1) theta@(Substitution m2 n2) =
   idempotencyCheck $

### 2018-12-01 paper/intro.tex
Tweak some of the typesetting
Start over on the paper

--- 2018-05-18 paper/intro.tex
@@
 
-In the two decades since they were introduced by \citet{wadler89}, \emph{type classes} have proven a versatile apparatus for structuring programs.
-Consider a class intended to characterize types $\tVar$ for which a monoid can be defined:
-\begin{flalign*}
-  & \eClass{\tMonoid{\tVar}} & \\
-  & \quad \tAnno{\eIdentity}{\tVar} & \\
-  & \quad \tAnno{\eCombine}{\tArrow{\tVar}{\tArrow{\tVar}{\tVar}}} &
-\end{flalign*}
-Programmers can use this class to define polymorphic functions over all monoids:
-\begin{flalign*}
-  & \tAnno{\eSquare}{\tMonoid{\tVar} \Rightarrow \tArrow{\tVar}{\tVar}} & \\
-  & \eApp{\eSquare}{\eX} = \eApp{\eApp{\eCombine}{\eX}}{\eX} &
-\end{flalign*}
-Finally, an \emph{instance} of the class can be provided for a specific type:
-\begin{flalign*}
-  & \eInstance{\tMonoid{\integer}} & \\
-  & \quad \eIdentity = \numberName{1} & \\
-  & \quad \eCombine = \eMultiply &
-\end{flalign*}
-where $\tAnno{\numberName{1}}{\integer}$ and $\tAnno{\eMultiply}{\tArrow{\integer}{\tArrow{\integer}{\integer}}}$.
-The instance is a witness to justify the well-definedness of expressions like $\eApp{\eSquare}{\numberName{3}}$.
+We call a monad \(\left(T, \eta, \mu\right)\) in a category \(\mathbb{C}\) \emph{algebraic} if every natural transformation \(\varepsilon : T \rightarrow 1_\mathbb{C}\) makes the diagrams
 
-More recently, algebraic effects \citep{plotkin03} and handlers \citep{plotkin09} have become an attractive alternative to monads for modeling computational effects due to their compositionality.
+\medskip
+\noindent\begin{minipage}{\dimexpr.5\textwidth-.5\columnsep}
+  \centering
+  \begin{tikzcd}[row sep=huge, column sep=huge]
+    1_\mathbb{C} \arrow[r, "\eta"] \arrow[rd, "1"]
+      & T \arrow[d, "\varepsilon"] \\
+    & 1_\mathbb{C}
+  \end{tikzcd}
+\end{minipage}\begin{minipage}{\dimexpr.5\textwidth-.5\columnsep}
+  \centering
+  \begin{tikzcd}[row sep=huge, column sep=huge]
+    T^2
+        \arrow[r, "\mu"]
+        \arrow[d, "\varepsilon T"]
+      & T
+        \arrow[d, "\varepsilon"]
+        \arrow[dl, equal] \\
+    T
+        \arrow[r, "\varepsilon"]
+      & 1_\mathbb{C}
+  \end{tikzcd}
+\end{minipage}
+\medskip
+
+\noindent commute.
+If \(\left(T, \eta, \mu\right)\) and \(\left(U, \delta, \nu\right)\) are algebraic monads, then any pair of natural transformations \(\varepsilon : T \rightarrow 1_\mathbb{C}\) and \(\theta : U \rightarrow 1_\mathbb{C}\) induces a composite algebraic monad \(\left(T U, \eta U \circ \delta, \theta TU \circ \varepsilon UTU\right)\).
+Furthermore, \(\delta T \circ \eta \circ \theta \circ \varepsilon U : TU \rightarrow UT\) is a natural isomorphism, which is to say composition of algebraic monads is commutative.

### 2018-12-01 scripts/lint-tex.rb

@@ -40,6 +40,7 @@ PRELUDE_ARITIES = {
   'acmYear' => [1],
   'affiliation' => [1],
   'alpha' => [0],
+  'arrow' => [0],
   'author' => [1],
   'begin' => [1, 2],
   'beginAcm' => [0],
@@ -52,18 +53,22 @@ PRELUDE_ARITIES = {
   'boxed' => [1],
   'caption' => [1],
   'ccsdesc' => [1],
+  'centering' => [0],
   'circ' => [0],
   'citep' => [1],
   'citestyle' => [1],
   'citet' => [1],
   'city' => [1],
   'color' => [1],
+  'columnsep' => [0],
   'copyrightyear' => [1],
   'country' => [1],
   'cup' => [0],
   'date' => [1],
   'definecolor' => [3],
+  'delta' => [0],
   'diamond' => [0],
+  'dimexpr' => [0],
   'displaystyle' => [0],
   'documentclass' => [1],
   'else' => [0],
@@ -73,6 +78,7 @@ PRELUDE_ARITIES = {
   'endAcm' => [0],
   'endNonAcm' => [0],
   'enspace' => [0],
+  'eta' => [0],
   'fbox' => [1],
   'fboxsep' => [0],
   'fi' => [0],
@@ -103,17 +109,20 @@ PRELUDE_ARITIES = {
   'makeatother' => [0],
   'maketitle' => [0],
   'mapsto' => [0],
+  'mathbb' => [1],
   'mathbf' => [1],
   'mathclose' => [1],
   'mathit' => [1],
   'mathopen' => [1],
   'medskip' => [0],
+  'mu' => [0],
   'newcommand' => [0],
   'newif' => [0],
   'newtheorem' => [2],
   'noindent' => [0],
   'notin' => [0],
   'nsubseteq' => [0],
+  'nu' => [0],
   'omega' => [0],
   'operatorname' => [1],
   'orcid' => [1],
@@ -144,6 +153,8 @@ PRELUDE_ARITIES = {
   'textit' => [1],
   'textsc' => [1],
   'textsf' => [1],
+  'textwidth' => [0],
+  'theta' => [0],
   'title' => [1],
   'ttfamily' => [0],
   'useacmtrue' => [0],
